---
title: "Drought"
description: "Tracks drought conditions across the United States using the Palmer Index, the SPEI, and the U.S. Drought Monitor."
categories: [Weather and Climate, Drought]
execute:
  eval: true
---

```{r}
#| label: setup
#| include: false
source(here::here("R", "common.R"))
source(here::here("R", "drought.R"))

meta     <- read_meta(REPO)
pdsi     <- read_indicator(REPO, "drought_pdsi_annual.csv")
spei     <- read_indicator(REPO, "drought_spei_five_year.csv")
division <- read_indicator(REPO, "drought_change_by_division.csv")
category <- read_indicator(REPO, "drought_land_area_by_category.csv")
```

## Figures

::::: {.panel-tabset}

### Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
figure_caption(meta, "drought_pdsi_annual.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data/drought_pdsi_annual.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data-raw/drought_fig-1.csv" download><i class="bi bi-filetype-csv"></i> EPA source data (CSV)</a>
</div>

```{r}
#| label: fig-1
#| echo: false
fig_1(pdsi)
```

This chart shows annual values of the Palmer Drought Severity Index,
averaged over the entire area of the contiguous 48 states. Positive values
represent wetter-than-average conditions, while negative values represent
drier-than-average conditions. A value between -2 and -3 indicates moderate
drought, -3 to -4 is severe drought, and -4 or below indicates extreme
drought. The thicker line is a nine-year weighted average.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(pdsi))
```

</details>

### Figure 2

```{r}
#| label: fig-2-caption
#| echo: false
#| output: asis
figure_caption(meta, "drought_spei_five_year.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data/drought_spei_five_year.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data-raw/drought_fig-2.csv" download><i class="bi bi-filetype-csv"></i> EPA source data (CSV)</a>
</div>

```{r}
#| label: fig-2
#| echo: false
fig_2(spei)
```

This graph shows annual values of the SPEI, averaged over the entire area
of the contiguous 48 states. The line in this graph represents an average
SPEI value for each year, based on conditions over the preceding five years
(five-year SPEI). Positive values represent wetter-than-average conditions,
while negative values represent drier-than-average conditions.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-2-table
#| echo: false
knitr::kable(fig_2_table(spei))
```

</details>

### Figure 3

::: {.callout-note appearance="simple" icon=false}
EPA's own Figure 3 is a choropleth map of the 344 individual NOAA climate
divisions, coloring each by its change in five-year SPEI. This site has no
source of climate-division boundary geometry, so this figure instead plots
every division as a single point, grouped by state (driest states at top)
and colored by direction of change. Every division's own value is
reproduced exactly as EPA published it; only the presentation differs. See
the [drought](https://github.com/climateindicators/drought) repository's
`data-raw/PROVENANCE.md` for more detail.
:::

```{r}
#| label: fig-3-caption
#| echo: false
#| output: asis
figure_caption(meta, "drought_change_by_division.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data/drought_change_by_division.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data-raw/drought_fig-3.csv" download><i class="bi bi-filetype-csv"></i> EPA source data (CSV)</a>
</div>

```{r}
#| label: fig-3
#| echo: false
fig_3(division)
```

This shows the total change in drought conditions across the contiguous 48
states, based on the long-term average rate of change in the five-year SPEI
from 1900 to 2023, for each of NOAA's 344 climate divisions. Orange points
represent increased moisture (a wetter trend); blue points represent
decreased moisture (a drier trend).

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-3-table
#| echo: false
knitr::kable(fig_3_table(division))
```

</details>

### Figure 4

```{r}
#| label: fig-4-caption
#| echo: false
#| output: asis
figure_caption(meta, "drought_land_area_by_category.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data/drought_land_area_by_category.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/drought/main/data-raw/drought_fig-4.csv" download><i class="bi bi-filetype-csv"></i> EPA source data (CSV)</a>
</div>

```{r}
#| label: fig-4
#| echo: false
fig_4(category)
```

This chart shows the percentage of U.S. lands classified under drought
conditions from 2000 through 2023, using the U.S. Drought Monitor
classification system (see the chart legend above). The data cover all 50
states plus Puerto Rico. Each category (D0 through D4) is the percentage of
land in exactly that category, not that category or worse, so the five
bands stack to the total percentage of land in any drought category.

Experts update the U.S. Drought Monitor weekly and produce maps that
illustrate current conditions as well as short- and longer-term trends.
Major participants include the National Oceanic and Atmospheric
Administration, the U.S. Department of Agriculture, and the National
Drought Mitigation Center. For a map of current drought conditions, visit
the Drought Monitor website at:
[https://droughtmonitor.unl.edu](https://droughtmonitor.unl.edu/).

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-4-table
#| echo: false
knitr::kable(fig_4_table(category))
```

</details>

:::::

## Key Points

- Average drought conditions across the nation have varied over time. The
  1930s and 1950s saw the most widespread droughts, while the last 50 years
  have generally been wetter than average (see Figures 1 and 2). Over the
  entire period shown in Figures 1 and 2, the overall trend has been toward
  wetter conditions, with the SPEI increasing at a rate of about 0.03 units
  per decade (see Figure 2).

- Large and consistent decreases in the SPEI have been observed throughout
  the western United States. Decreases have been especially prominent in
  southwestern states such as California, Arizona, and New Mexico. The
  eastern United States—in particular the Midwest and Northeast—has
  experienced generally wetter conditions (see Figure 3).

- In terms of scale and duration, the droughts of the 1930s Dust Bowl era
  remain the most extreme in the historical record (see Figures 1 and 2).

- Over the period from 2000 through 2023, roughly 10 to 70 percent of the
  U.S. land area experienced conditions that were at least abnormally dry
  at any given time (see Figure 4). The years 2002–2003 and 2012–2013 had a
  relatively large area with at least abnormally dry conditions, while
  2009–2011, 2016–2017, and 2019 had substantially less area experiencing
  drought.

- During the latter half of 2012, more than half of the U.S. land area was
  covered by moderate or greater drought (see Figure 4). In several states,
  2012 was among the driest years on record.^[6](#ref-6)^ See Temperature
  and Drought in the Southwest for a closer look at recent drought
  conditions in one of the hardest-hit regions.

## Background

There are many definitions and types of drought. Meteorologists generally
define drought as a prolonged period of dry weather caused by a lack of
precipitation that results in a serious water shortage for some activity,
population, or ecological system. Drought can also be thought of as an
extended imbalance between precipitation and evaporation.

As average temperatures have risen because of climate change, the Earth's
water cycle has sped up through an increase in the rate of evaporation from
the Earth's surface (including soil, lakes, and reservoirs) and
transpiration from plants. An increase in evapotranspiration makes more
water available in the air for precipitation, but contributes to drying
over some land areas, leaving less moisture in the soil. As the climate
continues to change, many historically wet areas are likely to experience
increased precipitation (see the U.S. and Global Precipitation indicator)
and increased risk of flooding (see the Heavy Precipitation indicator),
while historically dry areas are likely to experience less precipitation
and increased risk of drought.^[1](#ref-1)^ As a result, since the 1950s,
some regions of the world have experienced an increase in some types of
drought, including western North America, southern Europe, and much of
Africa.^[2](#ref-2)^

Drought conditions can negatively affect agriculture, water supplies,
energy production, human health, and many other aspects of society. The
impacts vary depending on the type, location, intensity, and duration of
the drought. For example, effects on agriculture can range from slowed
plant growth to severe crop losses, while water supply impacts can range
from lowered reservoir levels and dried-up streams to major water
shortages. Prolonged droughts pose a particular threat to indigenous
populations because of their economic and cultural dependence on land and
water supplies. Warming and drought can threaten medicinal and culturally
important plants and animals and can reduce water quality and availability,
making tribal populations particularly vulnerable to waterborne
illnesses.^[3](#ref-3)^ Lower streamflow and groundwater levels can also
harm ecosystems more broadly, by harming plants and animals and increasing
the risk of wildfires (see the Wildfires indicator).

## About the Indicator

During the 20th century, many indices were created to measure drought
severity by looking at precipitation, soil moisture, stream flow,
vegetation health, and other variables.^[4](#ref-4)^ Figure 1 shows annual
values of the most widely used index, the Palmer Drought Severity Index,
which is calculated from precipitation and temperature measurements at
weather stations. An index value of zero represents the average moisture
conditions observed between 1931 and 1990 at a given location. A positive
value means conditions are wetter than average, while a negative value is
drier than average. Index values from locations across the contiguous 48
states have been averaged together to produce the national values shown in
Figure 1.

Figures 2 and 3 show the Standardized Precipitation Evapotranspiration
Index (SPEI), which measures the combination of water supply
(precipitation) and atmospheric water demand (evapotranspiration, which is
based on temperature) to determine whether a certain area is experiencing
extreme drought, extreme moisture, or conditions in between. This
combination of inputs is useful because it allows the SPEI to consider how
droughts might affect agriculture and ecosystems, which depend on a balance
between water supply and demand. SPEI values between -1 and 1 are
considered near normal for a given area, whereas values below -1 signify
drought and values above 1 signify unusually moist conditions. Because
drought conditions fluctuate naturally, it is helpful to look at average
conditions over several years to explore how drought is connected to
long-term climate change.^[5](#ref-5)^ For this reason, this indicator
presents average SPEI values over consecutive five-year periods. For
example, the 2023 SPEI score for a particular area is actually the average
of 60 months of SPEI data, from July 2018 through June 2023.

For a more detailed perspective on recent trends, Figure 4 shows an index
called the Drought Monitor, which is based on several indices (including
Palmer), along with additional factors such as snow water content,
groundwater levels, reservoir storage, pasture/range conditions, and other
impacts. The Drought Monitor uses codes from D0 to D4 (see the chart legend
above Figure 4) to classify drought severity. This part of the indicator
covers all 50 states and Puerto Rico.

## About the Data

### Indicator Notes

Natural variability in the Earth's climate means that drought trends may
vary slightly when measured over different time periods. Using a multi-year
index as shown in Figures 2 and 3 (five-year SPEI) is relevant for
examining climate trends and potential connections to climate
change.^[5](#ref-5)^ Periods with extreme drought conditions (e.g., 1930s)
can influence the long-term trends calculated for this indicator.

The SPEI values used for this indicator are based on weighted PRISM data by
location.^[10](#ref-10)^ The underlying method for calculating SPEI does
not account for variables such as solar radiation, humidity, and wind speed
that can all influence drought conditions. As a result, the SPEI shown here
is more reflective of the influence of temperature. This is a limitation
and potential source of uncertainty in interpreting drought conditions.

The U.S. Drought Monitor (Figure 4) offers a closer look at the percentage
of the country that is affected by drought. The period of record for this
index is relatively short, however, and thus too short-lived to be used for
assessing long-term climate trends or exploring how recent observations
compare with historical patterns. With several decades of data collection,
future versions of this indicator should be able to paint a more complete
picture of trends over time.

Overall, this indicator gives a broad overview of drought conditions in the
United States. It is not intended to replace local information that might
describe conditions more precisely for a particular region. The national
averages in Figures 1 and 2 are particularly limited in this regard, as
averaging drought metrics over such a large area can obscure drought
extremes occurring at regional scales. Thus, these national graphs might
understate the degree to which droughts are becoming more severe in some
areas while other places receive more rain as a result of climate change.

### Data Sources

Data for Figure 1 were obtained from the National Oceanic and Atmospheric
Administration's National Centers for Environmental Information, which
maintains a large collection of climate data online at:
[www.ncei.noaa.gov/access/monitoring/climate-at-a-glance](https://www.ncei.noaa.gov/access/monitoring/climate-at-a-glance).

Data for Figures 2 and 3 came from the WestWide Drought Tracker, which is a
collaboration between the University of Idaho, the Western Regional Climate
Center, and the Desert Research Institute. These data are available at:
[https://wrcc.dri.edu/wwdt/batchdownload.php](https://wrcc.dri.edu/wwdt/batchdownload.php).
The measured index values are derived from Oregon State University's PRISM
Climate Mapping Program. Figures 2 and 3 were constructed using methods
described by Vicente-Serrano et al. (2010).^[11](#ref-11)^

Data for Figure 4 were provided by the National Drought Mitigation Center.
Historical data in table form are available at:
[https://droughtmonitor.unl.edu/DmData/DataTables.aspx](https://droughtmonitor.unl.edu/DmData/DataTables.aspx).

## Technical Documentation

```{r}
#| label: td-link
#| echo: false
#| output: asis
technical_documentation_link(meta)
```

## References

<ol class="references">
  <li id="ref-1">Marvel, K., Su, W., Delgado, R., Aarons, S., Chatterjee, A., Garcia, M. E., Hausfather, Z., Hayhoe, K., Hence, D. A., Jewett, E. B., Robel, A., Singh, D., Tripati, A., & Vose, R. S. (2023). Chapter 2: Climate trends. In USGCRP (U.S. Global Change Research Program), <em>Fifth National Climate Assessment</em>. [https://doi.org/10.7930/NCA5.2023.CH2](https://doi.org/10.7930/NCA5.2023.CH2)</li>
  <li id="ref-2">IPCC (Intergovernmental Panel on Climate Change). (2021). <em>Climate change 2021—The physical science basis: Working Group I contribution to the Sixth Assessment Report of the Intergovernmental Panel on Climate Change</em> (V. Masson-Delmotte, P. Zhai, A. Pirani, S. L. Connors, C. Péan, S. Berger, N. Caud, Y. Chen, L. Goldfarb, M. I. Gomis, M. Huang, K. Leitzell, E. Lonnoy, J. B. R. Matthews, T. K. Maycock, T. Waterfield, O. Yelekçi, R. Yu, & B. Zhou, Eds.). Cambridge University Press. [https://doi.org/10.1017/9781009157896](https://doi.org/10.1017/9781009157896)</li>
  <li id="ref-3">Gamble, J. L., Balbus, J., Berger, M., Bouye, K., Campbell, V., Chief, K., Conlon, K., Crimmins, A., Flanagan, B., Gonzalez-Maddux, C., Hallisey, E., Hutchins, S., Jantarasami, L., Khoury, S., Kiefer, M., Kolling, J., Lynn, K., Manangan, A., McDonald, M., … Wolkin, A. F. (2016). Chapter 9: Populations of concern. In USGCRP (U.S. Global Change Research Program), <em>The impacts of climate change on human health in the United States: A scientific assessment</em> (pp. 247–286). [https://doi.org/10.7930/J0Q81B0T](https://doi.org/10.7930/J0Q81B0T)</li>
  <li id="ref-4">Heim, R. R. (2002). A review of twentieth-century drought indices used in the United States. <em>Bulletin of the American Meteorological Society</em>, <em>83</em>(8), 1149–1166. [https://doi.org/10.1175/1520-0477-83.8.1149](https://doi.org/10.1175/1520-0477-83.8.1149)</li>
  <li id="ref-5">Abatzoglou, J. T., McEvoy, D. J., & Redmond, K. T. (2017). The West Wide Drought Tracker: Drought monitoring at fine spatial scales. <em>Bulletin of the American Meteorological Society</em>, <em>98</em>(9), 1815–1820. [https://doi.org/10.1175/BAMS-D-16-0193.1](https://doi.org/10.1175/BAMS-D-16-0193.1)</li>
  <li id="ref-6">NOAA (National Oceanic and Atmospheric Administration) National Centers for Environmental Information. (2013). <em>Monthly national climate report for December 2012</em>. [www.ncei.noaa.gov/access/monitoring/monthly-report/national/201212](https://www.ncei.noaa.gov/access/monitoring/monthly-report/national/201212)</li>
  <li id="ref-7">NOAA (National Oceanic and Atmospheric Administration) National Centers for Environmental Information. (2024). <em>Climate at a glance</em> [Data set]. Retrieved February 1, 2024, from [www.ncei.noaa.gov/access/monitoring/climate-at-a-glance](https://www.ncei.noaa.gov/access/monitoring/climate-at-a-glance)</li>
  <li id="ref-8">Western Regional Climate Center. (2024). <em>WestWide Drought Tracker</em>. Retrieved January 1, 2024, from [https://wrcc.dri.edu/wwdt](https://wrcc.dri.edu/wwdt)</li>
  <li id="ref-9">National Drought Mitigation Center. (2024). <em>U.S. drought monitor</em>. Retrieved February 1, 2024, from [https://droughtmonitor.unl.edu](https://droughtmonitor.unl.edu)</li>
  <li id="ref-10">Daly, C., Halbleib, M., Smith, J. I., Gibson, W. P., Doggett, M. K., Taylor, G. H., Curtis, J., & Pasteris, P. P. (2008). Physiographically sensitive mapping of climatological temperature and precipitation across the conterminous United States. <em>International Journal of Climatology</em>, <em>28</em>(15), 2031–2064. [https://doi.org/10.1002/joc.1688](https://doi.org/10.1002/joc.1688)</li>
  <li id="ref-11">Vicente-Serrano, S. M., Beguería, S., & López-Moreno, J. I. (2010). A multiscalar drought index sensitive to global warming: The standardized precipitation evapotranspiration index. <em>Journal of Climate</em>, <em>23</em>(7), 1696–1718. [https://doi.org/10.1175/2009JCLI2909.1](https://doi.org/10.1175/2009JCLI2909.1)</li>
</ol>

::: {.callout-note appearance="simple" icon=false}
Text on this page is EPA's own published wording (a U.S. Government work,
not subject to domestic copyright), extracted from EPA's source document and
verified paragraph by paragraph against the published page, with Figure 3's
description edited to describe this page's own presentation (a dot strip,
not EPA's choropleth map) rather than EPA's. Data, chart code, and the full
extraction pipeline are maintained in the
[drought](https://github.com/climateindicators/drought) repository, which
this page reads its data from directly.
:::
