---
title: "Global Greenhouse Gas Emissions"
description: "This indicator describes emissions of greenhouse gases worldwide."
categories: [Atmosphere, Emissions]
execute:
  eval: true
---

This indicator describes emissions of greenhouse gases worldwide.

```{r}
#| label: setup
#| include: false
source(here::here("R", "common.R"))
source(here::here("R", "global-greenhouse-gas-emissions.R"))

meta   <- read_meta(REPO)
gas    <- read_indicator(REPO, "ghg_by_gas.csv")
sector <- read_indicator(REPO, "ghg_by_sector.csv")
region <- read_indicator(REPO, "co2_by_region.csv")
```

## Figures

::::: {.panel-tabset}

### Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
figure_caption(meta, "ghg_by_gas.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/global-greenhouse-gas-emissions/main/data/ghg_by_gas.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/global-greenhouse-gas-emissions/main/data-raw/global-ghg-emissions_figure%201_11-22-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-1
#| echo: false
fig_1(gas)
```

This figure shows worldwide emissions of carbon dioxide, methane, nitrous
oxide, and several fluorinated gases from 1990 to 2015. For consistency,
emissions are expressed in million metric tons of carbon dioxide equivalents.
These totals include emissions and sinks due to land-use change and forestry.

_HFCs are hydrofluorocarbons, PFCs are perfluorocarbons, and SF6 is sulfur
hexafluoride._

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(gas), digits = 2)
```

</details>

### Figure 2

```{r}
#| label: fig-2-caption
#| echo: false
#| output: asis
figure_caption(meta, "ghg_by_sector.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/global-greenhouse-gas-emissions/main/data/ghg_by_sector.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/global-greenhouse-gas-emissions/main/data-raw/global-ghg-emissions_figure%202_11-22-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-2
#| echo: false
fig_2(sector)
```

This figure shows worldwide greenhouse gas emissions by sector from 1990 to
2015. For consistency, emissions are expressed in million metric tons of
carbon dioxide equivalents. These totals include emissions and sinks due to
land-use change and forestry.

Note that the sectors shown here are different from the economic sectors used
in U.S. emissions accounting (see the U.S. Greenhouse Gas Emissions
indicator). Emissions from international transport (aviation and marine) are
separate from the energy sector because they are not part of individual
countries' emissions inventories. The energy sector includes all other
transportation activities.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-2-table
#| echo: false
knitr::kable(fig_2_table(sector), digits = 2)
```

</details>

### Figure 3

```{r}
#| label: fig-3-caption
#| echo: false
#| output: asis
figure_caption(meta, "co2_by_region.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/global-greenhouse-gas-emissions/main/data/co2_by_region.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/global-greenhouse-gas-emissions/main/data-raw/global-ghg-emissions_figure%203_11-22-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-3
#| echo: false
fig_3(region)
```

This figure shows carbon dioxide emissions from 1990 to 2021 for different
regions of the world. These totals do not include emissions or sinks related
to land-use change or forestry. Inclusion of land-use change and forestry
would increase the apparent emissions from some regions while decreasing the
emissions from others.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-3-table
#| echo: false
knitr::kable(fig_3_table(region), digits = 2)
```

</details>

:::::

## Key Points

- In 2015, estimated worldwide emissions from human activities totaled nearly
  47 billion metric tons of greenhouse gases, expressed as carbon dioxide
  equivalents. This represents a 44 percent increase from 1990 (see Figures 1
  and 2). These numbers represent net emissions, which include the effects of
  land use and forestry.

- Between 1990 and 2015, global emissions of all major greenhouse gases
  increased (see Figure 1). Net emissions of carbon dioxide increased by 50
  percent, which is particularly important because carbon dioxide accounts
  for about three-fourths of total global emissions. Methane emissions
  increased the least—19 percent—while emissions of nitrous oxide increased
  by 26 percent. Emissions of fluorinated gases more than tripled.

- Energy production and use (including fuels used by vehicles and buildings)
  represent the largest source of greenhouse gas emissions worldwide (about
  74 percent of the total in 2015), followed by agriculture (12 percent in
  2015) (see Figure 2). While land-use change and forestry represent a net
  sink for emissions in the United States, absorbing carbon dioxide and
  offsetting emissions from other sources (see the U.S. Greenhouse Gas
  Emissions indicator), these activities are a net source of emissions on a
  global scale, largely because of deforestation.^[1](#ref-1)^

- Carbon dioxide emissions are increasing faster in some parts of the world
  (for example, East Asia and the Pacific) than in others (see Figure 3). The
  majority of emissions come from three regions: East Asia and Pacific,
  Europe and Central Asia, and the United States, which together accounted
  for 75 percent of total global emissions in 2021. The apparent downturn in
  2020 likely reflects the worldwide reduction in travel and economic
  activity during the COVID-19 pandemic, as reported in the U.S. Greenhouse
  Gas Emissions indicator.

## Background

Increasing emissions of greenhouse gases due to human activities worldwide,
especially burning of fossil fuels, have led to a substantial increase in
atmospheric concentrations of long-lived and other greenhouse gases (see the
Atmospheric Concentrations of Greenhouse Gases indicator). Every country
around the world emits greenhouse gases into the atmosphere, meaning the root
cause of climate change is truly global in scope. Some countries produce far
more greenhouse gases than others, and several factors—such as economic
activity (including the composition and efficiency of the economy),
population, income level, land use, and climatic conditions—can influence a
country's emissions levels. Tracking greenhouse gas emissions worldwide
provides a global context for understanding the United States' and other
nations' roles in climate change.

## About the Indicator

Like the U.S. Greenhouse Gas Emissions indicator, this indicator focuses on
emissions of gases covered under the United Nations Framework Convention on
Climate Change: carbon dioxide, methane, nitrous oxide, and several
fluorinated gases. These are all important greenhouse gases that are
influenced by human activities, and the Convention requires participating
countries to develop and periodically submit an inventory of these emissions.

Data and analysis for this indicator come from Climate Watch, which is
managed by the World Resources Institute. Climate Watch compiles data from
peer-reviewed and internationally recognized greenhouse gas inventories
developed by EPA and other government agencies worldwide. Global estimates
for carbon dioxide are published annually, but estimates for other gases,
such as methane and nitrous oxide, are available only every fifth year.
Climate Watch includes estimates of emissions and sinks associated with land
use and forestry activities, which come from global estimates compiled by the
Food and Agriculture Organization of the United Nations.

Each greenhouse gas has a different lifetime (how long it stays in the
atmosphere) and a different ability to trap heat in our atmosphere. To allow
different gases to be compared and added together, emissions are converted
into carbon dioxide equivalents. This step uses each gas's 100-year global
warming potential, which measures how much a given amount of the gas is
estimated to contribute to global warming over a period of 100 years after
being emitted. Carbon dioxide is assigned a global warming potential equal to
1. This analysis uses global warming potentials from the Intergovernmental
Panel on Climate Change's (IPCC's) Fourth Assessment Report. In that report,
methane has a global warming potential of 25, which means a ton of methane
emissions contributes 25 times as much warming as a ton of carbon dioxide
emissions over 100 years, and that ton of methane emissions is therefore
equal to 25 tons of carbon dioxide equivalents. See [the
table](https://www.epa.gov/climate-indicators/greenhouse-gases) for global
warming potentials from IPCC's Sixth Assessment Report.

## About the Data

### Indicator Notes

Like the U.S. Greenhouse Gas Emissions indicator, this indicator does not
include emissions of gases that affect climate but are not covered under the
United Nations Framework Convention on Climate Change. For example, this
indicator excludes ozone-depleting substances such as chlorofluorocarbons
(CFCs) and hydrochlorofluorocarbons (HCFCs), which have high global warming
potentials, because these gases have been or are currently being phased out
under an international agreement called the Montreal Protocol. This
indicator also excludes black carbon and aerosols, which most emissions
inventories do not cover. There are also various emissions of greenhouse
gases of natural origin, which this indicator does not cover.

Global emissions inventories for gases other than carbon dioxide are limited
to five-year intervals. The United Nations Framework Convention on Climate
Change database has more comprehensive data; however, these data are
available mainly for a group of mostly developed countries that account for
only about half of global greenhouse gas emissions. Thus, to provide a more
representative measure of global greenhouse gas emissions, this indicator
uses the broader Climate Watch database.

### Data Sources

Data for this indicator came from the World Resources Institute's Climate
Watch database, which is accessible online at:
[www.climatewatchdata.org/ghg-emissions](http://www.climatewatchdata.org/ghg-emissions).
The Climate Watch dataset consists of data that were originally collected by
organizations including the International Energy Agency, the Food and
Agriculture Organization of the United Nations, EPA, and the Global Carbon
Project, as well as Robbie M. Andrew (CICERO Center for International
Climate Research). Other global emissions estimates—such as the estimates
published by the Intergovernmental Panel on Climate Change^[1](#ref-1)^—are
based on many of the same sources.

## Technical Documentation

```{r}
#| label: td-link
#| echo: false
#| output: asis
technical_documentation_link(meta)
```

## References

<ol class="references">
  <li id="ref-1">IPCC (Intergovernmental Panel on Climate Change). (2023). <em>Climate change 2022—Mitigation of climate change: Working Group III contribution to the Sixth Assessment Report of the Intergovernmental Panel on Climate Change</em> (P. R. Shukla, J. Skea, R. Slade, R. Fradera, M. Pathak, A. A. Khourdajie, M. Belkacemi, R. van Diemen, A. Hasija, G. Lisboa, S. Luz, J. Malley, D. McCollum, S. Some, & P. Vyas, Eds.). Cambridge University Press. [https://doi.org/10.1017/9781009157926](https://doi.org/10.1017/9781009157926)</li>
  <li id="ref-2">Climate Watch. (2024). <em>Historical GHG emissions</em>. Retrieved November 22, 2024, from [www.climatewatchdata.org/ghg-emissions](http://www.climatewatchdata.org/ghg-emissions)</li>
</ol>

::: {.callout-note appearance="simple" icon=false}
Text on this page is EPA's own published wording (a U.S. Government work, not
subject to domestic copyright), extracted from EPA's source document and
verified paragraph by paragraph against the published page. Data, chart code,
and the full extraction pipeline are maintained in the
[global-greenhouse-gas-emissions](https://github.com/climateindicators/global-greenhouse-gas-emissions)
repository, which this page reads its data from directly.
:::
