---
title: "A Closer Look: Heat-Related Workplace Deaths"
description: "Deaths of U.S. workers attributed to heat exposure on the job."
categories: [Health, Temperature]
execute:
  eval: true
---

This feature presents data on worker deaths classified as "heat-related" in the United States and explores factors that influence exposure and vulnerability to heat.

```{r}
#| label: setup
#| include: false
source(here::here("R", "common.R"))
source(here::here("R", "heat-related-workplace-deaths.R"))

meta   <- read_meta(REPO)
annual <- read_indicator(REPO, "workplace_heat_deaths_annual.csv")
```

## Figures

::::: {.panel-tabset}

### Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
figure_caption(meta, "workplace_heat_deaths_annual.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heat-related-workplace-deaths/main/data/workplace_heat_deaths_annual.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heat-related-workplace-deaths/main/data-raw/heat-related%20workplace%20deaths%20figure%201%2004-30-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-1
#| echo: false
fig_1(annual)
```

This graph shows the number of reported occupational heat-related deaths in
the United States each year. The orange portions of the bars show deaths in
the construction sector and the blue portions show deaths in all other
sectors of the economy. Total construction sector deaths in 2019 fell below
the Bureau of Labor Statistics' reporting threshold; for that year, BLS
combined construction into a single "all-industry" total rather than
reporting a sector total (shown in green above). BLS has indicated that data
collection for 2020 fell below reporting thresholds (no bar that year).

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(annual), digits = 0)
```

</details>

### Example 1

```{r}
#| label: example-1-caption
#| echo: false
#| output: asis
figure_caption(meta, "workplace_heat_deaths_outdoor_workers_county.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heat-related-workplace-deaths/main/images/epa-example-1-outdoor-workers-map.png" download><i class="bi bi-filetype-png"></i> Figure image (PNG)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heat-related-workplace-deaths/main/data/workplace_heat_deaths_outdoor_workers_county.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heat-related-workplace-deaths/main/data-raw/heat-related%20workplace%20deaths%20example%201%2004-30-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

![Example 1. Prevalence of Outdoor Workers in the United States,
2018–2022](https://raw.githubusercontent.com/climateindicators/heat-related-workplace-deaths/main/images/epa-example-1-outdoor-workers-map.png){fig-alt="County-level map of the United States showing the percentage of each county's workforce in predominantly-outdoor occupations, ranging from under 4 percent (pale yellow) to over 40 percent (dark red), concentrated in parts of the West, the northern and southern Great Plains, and Alaska." fig-align="center"}

This map shows the percentage of each county's workforce engaged in
occupations that typically involve a substantial amount of time outdoors.
Job categories included in this analysis include delivery services, law
enforcement and emergency services, utilities, mining/quarrying/oil and gas
extraction, agriculture, fishing, forestry, and construction. The map shows
a five-year average.

This is EPA's own published map image. The county-level data behind it—one
row per county, five-year average percentage—is maintained in the
[heat-related-workplace-deaths](https://github.com/climateindicators/heat-related-workplace-deaths)
repository and available for download above; an interactive version built
from that data is not part of this indicator yet. This figure's table is
3,274 rows (one per county), so it is offered as a download rather than
rendered inline.

:::::

## Key Points

- From 1992 to 2022, a total of 986 workers across all industry sectors in
  the United States died from exposure to heat. This represents an average
  of 34 deaths per year, according to reported data (Figure 1).
- In the same period, 334 construction workers lost their lives to heat
  exposure on the job—meaning that the construction sector accounted for
  about 34 percent of all occupational heat-related deaths (Figure 1).
- The total number of heat-related worker deaths per year has ranged from
  11 (reported in 1992) to 61 (2011). The number of heat-related deaths in
  the construction sector peaked at 18 deaths in 2010 and 2011 (Figure 1).
- It is difficult to say for sure whether occupational heat-related deaths
  have increased or decreased meaningfully since 1992. This is because of
  high year-to-year variability and the possibility that the full extent of
  the problem is larger than the numbers reported here. Some deaths might
  not be properly categorized as heat-related, and there are reasons to
  believe that some work-related deaths might go unreported by employers,
  particularly in industries with many undocumented workers (see the
  "Notes" section below).

## Background

Extreme heat is the leading weather-related killer in the United States
(see the [Heat-Related Deaths indicator](heat-related-deaths.qmd)).^[1](#ref-1)^
Over the last few decades, unusually hot summer days and nights have become
more common across the contiguous 48 states (see the [High and Low
Temperatures
indicator](https://19january2025snapshot.epa.gov/climate-indicators/climate-change-indicators-high-and-low-temperatures))
and it is expected that extreme heat events will occur with greater
frequency and intensity in the future.^[2](#ref-2)^ Thus, heat-related
deaths are also expected to increase,^[3](#ref-3)^ although these increases
may be tempered as people physiologically adapt to higher temperatures and
as communities increase their resilience to heat.

Outdoor workers, particularly those engaging in strenuous physical
activity, disproportionately face greater heat-related health
threats.^[4](#ref-4)^ For example, construction workers represent only 6
percent of the total U.S. workforce, yet they account for more than
one-third of all reported occupational deaths associated with heat
exposure.^[5](#ref-5)^ Farmworkers also have disproportionately high rates
of heat-related illnesses and deaths.^[6](#ref-6)^ Worker safety and
productivity will likely be increasingly challenged as temperatures
outdoors continue to rise.^[3](#ref-3)^ To reduce risks, employers can
implement preventative measures such as ensuring access to water, training
workers to recognize signs of heat-related illnesses, mandating frequent
breaks and shade, providing heat-appropriate protective equipment, and
limiting work during the hottest times of day.^[5](#ref-5)^ Several states
have enacted rules to mandate these types of practices, as tracked by
organizations like the [National Caucus of Environmental
Legislators](https://www.ncelenviro.org/articles/state-action-on-extreme-heat-and-worker-protections/).
For more information, see the U.S. Occupational Safety and Health
Administration's (OSHA's) worker safety website:
[www.osha.gov/heat](https://www.osha.gov/heat) and:
[www.heat.gov/pages/who-is-at-risk-to-extreme-heat](https://www.heat.gov/pages/who-is-at-risk-to-extreme-heat).

## About the Feature

This feature tracks the number of workers who died as a result of heat
exposure on the job in the United States from 1992 to 2022. These data are
compiled by the federal Bureau of Labor Statistics (BLS) from multiple
sources. Many of these cases are tracked by OSHA, which requires all
employers under its jurisdiction to report any work-related death.
Additional information comes from other federal agencies (including the
military), state agencies, death certificates and autopsy reports, and
other corroborating evidence. The cause of death and other details about
each fatal incident are transcribed and coded using standardized
classification systems.

This feature focuses on cases whose "source of injury" was coded as
"Heat—environmental": that is, deaths caused by high temperatures that
occur indoors or outdoors. Figure 1 shows the total number of heat-related
deaths per year. It also shows the number of deaths specific to the
construction sector, which is consistently the sector with the most
heat-related deaths.

## Exploring Worker Exposure and Vulnerability to Heat

Highlighting the intersections of climate hazards, such as extreme heat,
and social vulnerability can help identify people and places most at risk.
Outdoor workers, particularly those engaging in strenuous physical
activity, disproportionately face heat-related health threats. Between 5
million and 10 million American workers are exposed to outdoor heat beyond
safe levels every year.^[8](#ref-8)^ Outdoor occupations are distributed
throughout the United States, and they make up an especially large
percentage of the workforce in certain regions (see Example 1 above),
including parts
of the West, northern and southern Great Plains, and Alaska. Outdoor
workers in cities may also be exposed to elevated temperatures due to the
heat island effect, which can increase temperatures by 18 to 27°F during
the day and 9 to 18°F at night in urban areas compared with outlying
areas.^[9](#ref-9)^

## Notes

This feature counts deaths in workplaces of all types and sizes. It is not
limited to establishments that fall under the jurisdiction of OSHA or other
specific agencies.

Several factors influence the ability of this feature to estimate the true
number of occupational heat-related deaths. As discussed in the
[Heat-Related Deaths indicator](heat-related-deaths.qmd),
deaths classified as heat-related may have other important contributing
causes. Conversely, deaths associated with extreme heat are not always
reported as such. Unlike the Heat-Related Deaths indicator, though, this
feature is based on many types of corroborating evidence in addition to
death certificates. A fraction of deaths may go unreported by employers,
may not be properly categorized as heat-related, or may not be attributed
to an occupational cause if the death occurs off the job site (for example,
after the employee has left work for the day).^[11](#ref-11)^ The
underreporting rate is believed to be larger in industries that tend to
have more undocumented or short-term workers.^[12](#ref-12)^

This feature uses death counts rather than death rates. Reporting the total
number of deaths per year is useful for communication because it puts the
exact number of lives in perspective, but it does not account for changes
over time in population or sector-specific employment. Improvements to the
tracking of worker-related fatalities include standardization of employer
and state reporting of work-related heat deaths.^[13](#ref-13)^ Reporting
of illnesses and injuries must be immediate to ensure that illnesses and
deaths are not mis-classified or missed altogether.^[14](#ref-14)^

## Data Sources

Data for this feature were provided by BLS. The data in Figure 1 are
publicly available on the BLS Injuries, Illnesses, and Fatalities website
at: [www.bls.gov/iif/home.htm](https://www.bls.gov/iif/home.htm). Data for
the outdoor workers map figure came from BLS's Quarterly Census of
Employment and Wages:
[www.bls.gov/cew/downloadable-data-files.htm](https://www.bls.gov/cew/downloadable-data-files.htm).

## Technical Documentation

```{r}
#| label: td-link
#| echo: false
#| output: asis
technical_documentation_link(meta)
```

## References

<ol class="references">
  <li id="ref-1">NOAA (National Oceanic and Atmospheric Administration). (2023). _Weather related fatality and injury statistics_. Retrieved May 1, 2024, from www.weather.gov/hazstat</li>
  <li id="ref-2">Marvel, K., Su, W., Delgado, R., Aarons, S., Chatterjee, A., Garcia, M. E., Hausfather, Z., Hayhoe, K., Hence, D. A., Jewett, E. B., Robel, A., Singh, D., Tripati, A., & Vose, R. S. (2023). Chapter 2: Climate trends. In USGCRP (U.S. Global Change Research Program), _Fifth National Climate Assessment_. https://doi.org/10.7930/NCA5.2023.CH2</li>
  <li id="ref-3">IPCC (Intergovernmental Panel on Climate Change). (2022). _Climate change 2022—Impacts, adaptation and vulnerability: Working Group II contribution to the Sixth Assessment Report of the Intergovernmental Panel on Climate Change_ (H.-O. Pörtner, D. C. Roberts, M. Tignor, E. S. Poloczanska, K. Mintenbeck, A. Alegría, M. Craig, S. Langsdorf, S. Löschke, V. Möller, A. Okem, & B. Rama, Eds.). Cambridge University Press. https://doi.org/10.1017/9781009325844</li>
  <li id="ref-4">Hoegh-Guldberg, O., Jacob, D., Taylor, M., Bindi, M., Brown, S., Camilloni, I., Diedhiou, A., Djalante, R., Ebi, K. L., Engelbrecht, F., Guiot, J., Hijioka, Y., Mehrotra, S., Payne, A., Seneviratne, S. I., Thomas, A., Warren, R., & Zhou, G. (2018). Chapter 3: Impacts of 1.5oC global warming on natural and human systems. In IPCC (Intergovernmental Panel on Climate Change), _Global warming of 1.5°C. An IPCC special report on the impacts of global warming of 1.5°C above pre-industrial levels and related global greenhouse gas emission pathways, in the context of strengthening the global response to the threat of climate change, sustainable development, and efforts to eradicate poverty_ (pp. 175–312). Cambridge University Press. www.ipcc.ch/sr15/chapter/chapter-3</li>
  <li id="ref-5">Dong, X. S., West, G. H., Holloway-Beth, A., Wang, X., & Sokas, R. K. (2019). Heat-related deaths among construction workers in the United States. _American Journal of Industrial Medicine_, _62_(12), 1047–1057. https://doi.org/10.1002/ajim.23024</li>
  <li id="ref-6">Jay, A. K., Crimmins, A. R., Avery, C. W., Dahl, T. A., Dodder, R. S., Hamlington, B. D., Lustig, A., Marvel, K., Méndez-Lazaro, P. A., Osler, M. S., Terando, A., Weeks, E. S., & Zycherman, A. (2023). Chapter 1: Overview: Understanding risks, impacts, and responses. In USGCRP (U.S. Global Change Research Program), _Fifth National Climate Assessment_. https://doi.org/10.7930/NCA5.2023.CH1</li>
  <li id="ref-7">BLS (U.S. Bureau of Labor Statistics). (2024). _Census of Fatal Occupational Injuries (CFOI)_. Retrieved September 10, 2024, from www.bls.gov/iif/home.htm.</li>
  <li id="ref-8">Milken Institute School of Public Health. (2017). _Hazard zone: The impact of climate change on occupational health_. Retrieved April 29, 2024, from https://onlinepublichealth.gwu.edu/resources/impact-of-climate-change-on-occupational-health</li>
  <li id="ref-9">National Integrated Heat Health Information System. (2024). _Who is most at risk to extreme heat?_ Retrieved April 1, 2024, from www.heat.gov/pages/who-is-at-risk-to-extreme-heat</li>
  <li id="ref-10">BLS (U.S. Bureau of Labor Statistics). (2024). _Quarterly Census of Employment and Wages (QCEW)_ [Data set]. Retrieved April 1, 2024, from www.bls.gov/cew/downloadable-data-files.htm</li>
  <li id="ref-11">Hayden, M. H., Schramm, P. J., Beard, C. B., Bell, J. E., Bernstein, A. S., Bieniek-Tobasco, A., Cooley, N., Diuk-Wasser, M., Dorsey, M. K., Ebi, K., Ernst, K. C., Gorris, M. E., Howe, P. D., Khan, A. S., Lefthand-Begay, C., Maldonado, J., Saha, S., Shafiei, F., Vaidyanathan, A., & Wilhelmi, O. V. (2023). Chapter 15: Human health. In USGCRP (U.S. Global Change Research Program), _Fifth National Climate Assessment_. https://doi.org/10.7930/NCA5.2023.CH15</li>
  <li id="ref-12">The White House. (2021). _Fact sheet: Biden administration mobilizes to protect workers and communities from extreme heat_. www.whitehouse.gov/briefing-room/statements-releases/2021/09/20/fact-sheet-biden-administration-mobilizes-to-protect-workers-and-communities-from-extreme-heat</li>
  <li id="ref-13">OSHA (Occupational Safety and Health Administration). (2022). _Heat: Standards_. Retrieved November 1, 2022, from www.osha.gov/heat-exposure/standards</li>
  <li id="ref-14">Orrenius, P. M., & Zavodny, M. (2009). Do immigrants work in riskier jobs? _Demography_, _46_(3), 535–551. https://doi.org/10.1353/dem.0.0064</li>
</ol>

::: {.callout-note appearance="simple" icon=false}
Text on this page is EPA's own published wording (a U.S. Government work, not
subject to domestic copyright), extracted from EPA's source document and
verified paragraph by paragraph against the published page. Data, chart code,
and the full extraction pipeline are maintained in the
[heat-related-workplace-deaths](https://github.com/climateindicators/heat-related-workplace-deaths)
repository, which this page reads its data from directly.
:::
