---
title: "Heavy Precipitation"
description: "How the frequency and intensity of heavy precipitation events have changed across the United States."
categories: [Weather and Climate, Precipitation]
execute:
  eval: true
---

This indicator tracks the frequency of heavy precipitation events in the
United States.

```{r}
#| label: setup
#| include: false
source(here::here("R", "common.R"))
source(here::here("R", "heavy-precipitation.R"))

meta        <- read_meta(REPO)
oneday      <- read_indicator(REPO, "extreme_oneday_precipitation.csv")
annual_high <- read_indicator(REPO, "unusually_high_annual_precipitation.csv")
```

## Figures

::::: {.panel-tabset}

### Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
figure_caption(meta, "extreme_oneday_precipitation.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heavy-precipitation/main/data/extreme_oneday_precipitation.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heavy-precipitation/main/data-raw/heavy-precip_fig-1.csv" download><i class="bi bi-filetype-csv"></i> EPA source data (CSV)</a>
</div>

```{r}
#| label: fig-1
#| echo: false
fig_1(oneday)
```

This figure shows the percentage of the land area of the contiguous 48
states where a much greater than normal portion of total annual
precipitation has come from extreme single-day precipitation events. The
blue bars show the annual value for each year from 1910 to 2023. The orange
line is a nine-year weighted moving average, which EPA's source file
provides directly rather than something this site calculates.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(oneday), digits = 3)
```

</details>

### Figure 2

```{r}
#| label: fig-2-caption
#| echo: false
#| output: asis
figure_caption(meta, "unusually_high_annual_precipitation.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heavy-precipitation/main/data/unusually_high_annual_precipitation.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/heavy-precipitation/main/data-raw/heavy-precip_fig-2.csv" download><i class="bi bi-filetype-csv"></i> EPA source data (CSV)</a>
</div>

```{r}
#| label: fig-2
#| echo: false
fig_2(annual_high)
```

This figure shows the percentage of the land area of the contiguous 48
states that experienced much greater than normal precipitation in any given
year, which means it scored 2.0 or above on the annual Standardized
Precipitation Index. The thin blue line shows the annual value for each
year from 1895 to 2023. The thicker orange line is a nine-year weighted
moving average that smooths out some of the year-to-year fluctuations.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-2-table
#| echo: false
knitr::kable(fig_2_table(annual_high), digits = 3)
```

</details>

:::::

## Key Points

- In recent years, a larger percentage of precipitation has come in the
  form of intense single-day events. Nine of the top 10 years for extreme
  one-day precipitation events have occurred since 1995 (see Figure 1).

- The prevalence of extreme single-day precipitation events remained fairly
  steady between 1910 and the 1980s, but has risen substantially since
  then. Over the entire period from 1910 to 2023, the portion of the
  country experiencing extreme single-day precipitation events increased at
  a rate of about half a percentage point per decade (see Figure 1).

- The percentage of land area experiencing much greater than normal yearly
  precipitation totals increased between 1895 and 2023. There has been much
  year-to-year variability, however. In some years there were no abnormally
  wet areas, while a few others had abnormally high precipitation totals
  over 10 percent or more of the contiguous 48 states' land area (see
  Figure 2). For example, 1941 was extremely wet in the West, while 1983
  was very wet nationwide.^[3](#ref-3)^

- Figures 1 and 2 are both consistent with other studies that have found an
  increase in heavy precipitation over timeframes ranging from single days
  to seasons to years.^[1](#ref-1)^ For more information on trends in
  overall precipitation levels, see the U.S. and Global Precipitation
  indicator.

## Background

"Heavy precipitation" refers to instances during which the amount of rain
or snow experienced in a location substantially exceeds what is normal.
What constitutes a period of heavy precipitation varies according to
location and season.

Climate change can affect the intensity and frequency of precipitation.
Warmer oceans increase the amount of water that evaporates into the air.
When more moisture-laden air moves over land or converges into a storm
system, it can produce more intense precipitation—for example, heavier
rain and snow storms.^[1](#ref-1)^ The potential impacts of heavy
precipitation include crop damage, soil erosion, and an increase in flood
risk due to heavy rains (see the River Flooding indicator)—which in turn
can lead to injuries, drownings, respiratory health impacts from exposure
to mold, and other flooding-related effects on health.^[2](#ref-2)^ In
addition, runoff from precipitation can impair water quality as pollutants
deposited on land wash into water bodies.

Heavy precipitation does not necessarily mean the total amount of
precipitation at a location has increased—just that precipitation is
occurring in more intense events. However, changes in the intensity of
precipitation, when combined with changes in the interval between
precipitation events, can also lead to changes in overall precipitation
totals.

## About the Indicator

Heavy precipitation events can be measured by tracking their frequency,
examining their return period (the chance that the event will be equaled
or exceeded in a given year), or directly measuring the amount of
precipitation in a certain period (for example, inches of rain falling in
a 24-hour period).

One way to track heavy precipitation is by calculating what percentage of
a particular location's total precipitation in a given year has come in
the form of extreme one-day events—or, in other words, what percentage of
precipitation is arriving in more concentrated periods of time. Figure 1
of this indicator looks at the prevalence of extreme single-day
precipitation events over time.

For added insight, this indicator also tracks the occurrence of unusually
high total yearly precipitation. It does so by looking at the Standardized
Precipitation Index (SPI), which compares actual yearly precipitation
totals with the range of precipitation totals that one would typically
expect at a specific location, based on historical data. If a location
experiences less precipitation than normal during a particular period, it
will receive a negative SPI score, while a period with more precipitation
than normal will receive a positive score. The more precipitation
(compared with normal), the higher the SPI score. The SPI is a useful way
to look at precipitation totals because it allows comparison of different
locations and different seasons on a standard scale. Figure 2 shows what
percentage of the total area of the contiguous 48 states had an annual SPI
score of 2.0 or above (well above normal) in any given year.

## About the Data

### Indicator Notes

Weather monitoring stations tend to be closer together in the eastern and
central states than in the western states. In areas with fewer monitoring
stations, heavy precipitation indicators are less likely to reflect local
conditions accurately.

### Data Sources

The data used for this indicator come from a large national network of
weather stations and were provided by the National Oceanic and Atmospheric
Administration's (NOAA's) National Centers for Environmental Information.
Figure 1 is based on Step #4 of NOAA's U.S. Climate Extremes Index; for
data and a description of the index, see:
[www.ncei.noaa.gov/access/monitoring/cei](https://www.ncei.noaa.gov/access/monitoring/cei).
Figure 2 is based on the U.S. SPI, which is shown in a variety of maps
available online at:
[www.ncei.noaa.gov/access/monitoring/monthly-report/drought](https://www.ncei.noaa.gov/access/monitoring/monthly-report/drought).
The data used to construct these maps are available at:
[www.ncei.noaa.gov/pub/data/cirs/climdiv](https://www.ncei.noaa.gov/pub/data/cirs/climdiv).

## Technical Documentation

```{r}
#| label: td-link
#| echo: false
#| output: asis
technical_documentation_link(meta)
```

## References

<ol class="references">
  <li id="ref-1">Marvel, K., Su, W., Delgado, R., Aarons, S., Chatterjee, A., Garcia, M. E., Hausfather, Z., Hayhoe, K., Hence, D. A., Jewett, E. B., Robel, A., Singh, D., Tripati, A., & Vose, R. S. (2023). Chapter 2: Climate trends. In USGCRP (U.S. Global Change Research Program), <em>Fifth National Climate Assessment</em>. [https://doi.org/10.7930/NCA5.2023.CH2](https://doi.org/10.7930/NCA5.2023.CH2)</li>
  <li id="ref-2">Bell, J. E., Herring, S. C., Jantarasami, L., Adrianopoli, C., Benedict, K., Conlon, K., Escobar, V., Hess, J., Luvall, J., Garcia-Pando, C. P., Quattrochi, D., Runkle, J., & Schreck, C. J., III. (2016). Chapter 4: Impacts of extreme events on human health. In USGCRP (U.S. Global Change Research Program), <em>The impacts of climate change on human health in the United States: A scientific assessment</em> (pp. 99–128). [https://health2016.globalchange.gov/extreme-events](https://health2016.globalchange.gov/extreme-events)</li>
  <li id="ref-3">NOAA (National Oceanic and Atmospheric Administration). (2012). <em>Personal communication: Analysis by Derek Arndt, April 2012</em>. National Climatic Data Center.</li>
  <li id="ref-4">NOAA (National Oceanic and Atmospheric Administration). (2024). <em>U.S. Climate Extremes Index</em>. Retrieved April 18, 2024, from [www.ncei.noaa.gov/access/monitoring/cei](https://www.ncei.noaa.gov/access/monitoring/cei)</li>
  <li id="ref-5">NOAA (National Oceanic and Atmospheric Administration). (2024). <em>Standardized Precipitation Index data files</em> [Data set]. Retrieved April 18, 2024, from [www.ncei.noaa.gov/pub/data/cirs/climdiv](https://www.ncei.noaa.gov/pub/data/cirs/climdiv)</li>
</ol>

::: {.callout-note appearance="simple" icon=false}
Text on this page is EPA's own published wording (a U.S. Government work,
not subject to domestic copyright), extracted from EPA's source document
and verified paragraph by paragraph against the published page. Data,
chart code, and the full extraction pipeline are maintained in the
[heavy-precipitation](https://github.com/climateindicators/heavy-precipitation)
repository, which this page reads its data from directly.
:::
