---
title: "Residential Energy Use"
description: "How much electricity and natural gas U.S. homes use for summer cooling and winter heating, alongside cooling and heating degree days."
categories: [Society, Energy]
execute:
  eval: true
---

This indicator examines trends related to home air conditioning and heating by tracking the amount of electricity used by U.S. homes in the summer and energy used in the winter.

```{r}
#| label: setup
#| include: false
source(here::here("R", "common.R"))
source(here::here("R", "residential-energy-use.R"))

meta            <- read_meta(REPO)
summer_annual   <- read_indicator(REPO, "residential_energy_summer_electricity.csv")
winter_annual   <- read_indicator(REPO, "residential_energy_winter_natural_gas.csv")
summer_monthly  <- read_indicator(REPO, "residential_energy_summer_electricity_monthly.csv")
winter_monthly  <- read_indicator(REPO, "residential_energy_winter_natural_gas_monthly.csv")
```

## Figures

::::: {.panel-tabset}

### Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
figure_caption(meta, "residential_energy_summer_electricity.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data/residential_energy_summer_electricity.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data-raw/residential%20energy%20use%20-%20all%20figures%20-%2012-19-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-1
#| echo: false
fig_1(summer_annual)
```

This graph shows the amount of electricity used by the average American during
the summer months (June, July, and August) of each year from 1973 to 2024. The
top panel shows average summer electricity use per capita, representing all 50
states plus D.C. For reference, the bottom panel shows the average number of
cooling degree days for the same months across the contiguous 48 states plus
D.C.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(summer_annual), digits = 1)
```

</details>

### Figure 2

```{r}
#| label: fig-2-caption
#| echo: false
#| output: asis
figure_caption(meta, "residential_energy_winter_natural_gas.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data/residential_energy_winter_natural_gas.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data-raw/residential%20energy%20use%20-%20all%20figures%20-%2012-19-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-2
#| echo: false
fig_2(winter_annual)
```

This graph shows the amount of natural gas used by the average American during
the winter months (December, January, and February) of each year from 1974 to
2024 (winter "1974" covers December 1973 through February 1974, and so on).
The top panel shows average winter natural gas use per capita, representing
all 50 states plus D.C. For reference, the bottom panel shows the average
number of heating degree days for the same months across the contiguous 48
states plus D.C.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-2-table
#| echo: false
knitr::kable(fig_2_table(winter_annual), digits = 1)
```

</details>

### Figure TD-1

::: {.callout-note appearance="simple" icon=false}
This figure is not part of EPA's published indicator page. It is EPA's own
supplementary Figure TD-1 from the technical documentation, included here
because the source workbook has the data and it shows the relationship
behind Figure 1's summer electricity series. See the
[residential-energy-use](https://github.com/climateindicators/residential-energy-use)
repository's `data-raw/PROVENANCE.md` for details.
:::

```{r}
#| label: fig-td1-caption
#| echo: false
#| output: asis
figure_caption(meta, "residential_energy_summer_electricity_monthly.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data/residential_energy_summer_electricity_monthly.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data-raw/residential%20energy%20use%20-%20all%20figures%20-%2012-19-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-td1
#| echo: false
fig_td1(summer_monthly)
```

Each point is one summer month (June, July, or August) from 1973 to 2024,
plotting that month's electricity use per capita against its cooling degree
days. The dashed line is a simple linear fit, shown to illustrate the
relationship, not as a prediction.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-td1-table
#| echo: false
knitr::kable(fig_td1_table(summer_monthly), digits = 1)
```

</details>

### Figure TD-2

::: {.callout-note appearance="simple" icon=false}
This figure is not part of EPA's published indicator page. It is EPA's own
supplementary Figure TD-2 from the technical documentation, included here
because the source workbook has the data and it shows the relationship
behind Figure 2's winter natural gas series. See the
[residential-energy-use](https://github.com/climateindicators/residential-energy-use)
repository's `data-raw/PROVENANCE.md` for details.
:::

```{r}
#| label: fig-td2-caption
#| echo: false
#| output: asis
figure_caption(meta, "residential_energy_winter_natural_gas_monthly.csv")
```

<div class="d-flex flex-wrap gap-2 mb-3">
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data/residential_energy_winter_natural_gas_monthly.csv" download><i class="bi bi-filetype-csv"></i> Figure data (CSV)</a>
<a class="btn btn-sm btn-outline-secondary" href="https://raw.githubusercontent.com/climateindicators/residential-energy-use/main/data-raw/residential%20energy%20use%20-%20all%20figures%20-%2012-19-24.xlsx" download><i class="bi bi-file-earmark-excel"></i> EPA source workbook (XLSX)</a>
</div>

```{r}
#| label: fig-td2
#| echo: false
fig_td2(winter_monthly)
```

Each point is one winter month (December, January, or February) from
winter 1974 to winter 2024, plotting that month's natural gas use per capita
against its heating degree days. The dashed line is a simple linear fit,
shown to illustrate the relationship, not as a prediction.

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-td2-table
#| echo: false
knitr::kable(fig_td2_table(winter_monthly), digits = 1)
```

</details>

:::::

## Key Points

- The amount of electricity used by the average American at home during the
  summer has nearly doubled since 1973, but it appears to have leveled off
  somewhat in recent years (see Figure 1).

- Year-to-year fluctuations in electricity use generally follow changes in
  cooling degree days (see Figure 1). Total cooling degree days have also
  increased since 1973, though they have not increased as dramatically as
  electricity use.

- The amount of natural gas used by the average American at home during the
  winter has decreased since 1974 (see Figure 2).

- Year-to-year fluctuations in natural gas use appear to mirror changes in
  heating degree days (see Figure 2). Total heating degree days have also
  decreased since 1974, though they have not decreased as dramatically as
  natural gas use.

## Background

Hot weather can be uncomfortable, and in extreme cases dangerous (see the Heat
Waves indicator). One way people respond to hot weather is by using air
conditioning. Air conditioning is the primary way to cool a home and has
become more widely used over time; as of 2020, it accounted for 19 percent of
the electricity that the average American household uses every
year.^[1](#ref-1)^ Conversely, cold weather requires people to use energy to
heat their homes, using electricity and fuels such as natural gas, propane,
and oil.

As climate change contributes to an increase in average temperatures and
unusually hot days, Americans are expected to use more energy—mostly
electricity for air conditioning.^[2](#ref-2)^ As a result, people will likely
have to spend more money on electricity for cooling.^[2](#ref-2)^ Increased
use of air conditioning could also lead to more greenhouse gas emissions and
further climate change, because more electricity must be generated to meet
this increased demand. As of 2023, about three-fifths of the electricity in
the United States was generated by burning fossil fuels such as coal or
natural gas.^[3](#ref-3)^ Burning fossil fuels for electricity is one of the
largest sources of heat-trapping greenhouse gas emissions in the United States
(see the U.S. Greenhouse Gas Emissions indicator).

The opposite effect could take place with winter heating. As temperatures
rise, Americans are expected to use less fuel energy for heating their
homes.^[2](#ref-2)^ Overall, though, summer increases in energy use are
expected to outweigh any reduction due to lower heating needs in the winter.^[2](#ref-2)^
Improvements in energy efficiency, particularly for electricity-based heating
and cooling, can influence residential consumption and demand for energy over
time.

## About the Indicator

This indicator measures changes in residential seasonal energy use in the
United States. It has two components:

First, it looks at the amount of electricity that U.S. homes use during the
summer (defined here as June, July, and August), when air conditioning is
typically needed most.

Next, it looks at the amount of natural gas that U.S. homes use during the
winter (defined here as December, January, and February), when heating is
typically needed most. Although a variety of heating fuels are used, natural
gas is the most representative for a national analysis because it accounts
for about 72 percent of the energy used for home heating in the United
States.^[1](#ref-1)^

The data come from the U.S. Energy Information Administration (EIA), which
collects comprehensive information about power generation, delivery, and
sales from electric power utilities across the country. EIA has collected
these data consistently since 1973. For this indicator, EPA divided total
residential electricity sales for the summer months by the corresponding
total U.S. population to calculate summer electricity use per capita, which
Figure 1 shows in kilowatt-hours per person. EPA used a similar method with
winter residential natural gas sales, which Figure 2 shows in cubic feet per
person.

For reference, Figure 1 also shows summer cooling degree days for each year,
while Figure 2 shows winter heating degree days for each year (see the
Heating and Cooling Degree Days indicator). In general, more cooling degree
days and fewer heating degree days means a combination of more warm days and
warmer temperatures on those days. Residential electricity and natural gas
use cover all 50 states plus Washington, D.C., while cooling and heating
degree days are shown for the contiguous 48 states plus D.C., as they are not
routinely calculated for Alaska and Hawaii.

## About the Data

### Indicator Notes

This indicator is based on residential retail electricity and natural gas
sales. In the case of electricity, sales are not exactly the same as total
residential electricity use, because sales data do not count electricity that
people generate and consume onsite—for example, using rooftop solar panels.
For context, residential solar generated an estimated 38,900 gigawatt-hours
of electricity in 2022, which translates to 2.6 percent of total residential
retail sales.^[7](#ref-7)^ As more people produce their own power and buy
less electricity from a utility, sales data could increasingly underestimate
the actual amount of electricity used. However, retail sales data are still
the best available approximation for energy use nationwide.

Outdoor air temperature is not the only weather-related factor that
influences energy use (humidity and dew point also matter, for example), but
it is still a major driver of the amount of electricity and natural gas
people use in their homes. That said, rising temperatures are not the only
reason why people might use more electricity in the summer and less natural
gas in the winter.

One possible factor affecting electricity and natural gas use is where people
live. Over the past several decades, the fastest-growing parts of the United
States have been in the South and West, which means a growing share of the
U.S. population lives in warmer places. The "cooling degree days" line in
Figure 1 and the "heating degree days" line in Figure 2 adjust for where
people lived as of the 2020 census but do not adjust for changes in
population distribution over time.

Other factors include characteristics of people's homes and the air
conditioning and heating equipment they use. As of 2020, 89 percent of U.S.
homes use some type of air conditioning, and 67 percent have central air
conditioning installed.^[1](#ref-1)^ These numbers have increased over time,
as has the size of the average American home, which means more space to heat
and cool. Conversely, homes have become more energy efficient over time (for
example, through better insulation), and air conditioning and heating
equipment has also become more efficient.

Electricity and natural gas use could also reflect utility prices (when rates
are higher, people may choose to reduce their use) and the amount of
electricity and natural gas that Americans use for other needs. Other
considerations include whether new homes rely on electricity or gas for
cooking, water heating, and other activities, and shifts towards the use of
innovative electronic appliances and devices over time.

Even with all these factors at work, many studies agree that rising
temperatures can and will lead to increased use of electricity, as well as
reduced use of heating fuels in winter.^[2](#ref-2)^

### Data Sources

The data for this indicator were provided by EIA, which maintains a large
collection of information about how the United States produces and uses
electricity and other forms of energy. EIA's energy sales data can be found
online at: [www.eia.gov/electricity](https://www.eia.gov/electricity/) and
[www.eia.gov/naturalgas](https://www.eia.gov/naturalgas). EPA developed this
indicator using two additional sources: population data from the U.S. Bureau
of Economic Analysis and cooling and heating degree days from the National
Oceanic and Atmospheric Administration.

## Statistical/Trend Analysis

To test for the presence of long-term national-level changes, the residential
summer electricity use per capita and residential winter natural gas use per
capita data series in Figures 1 and 2, respectively, were analyzed with an
ordinary least squares linear regression of annual data points. For
electricity, this results in a trend of +13.9 kilowatt-hours per year. This
trend is statistically significant (p < 0.001). For natural gas, this
results in a trend of -61.7 cubic feet per year. This trend, too, is
statistically significant (p < 0.001).

EPA has also tested these trends using a Sen's slope regression, which is a
non-parametric approach that finds the median of all possible pairwise slopes
in a temporal data set (Sen, 1968; Theil, 1950). The results are similar:
+14.4 kilowatt-hours per year for summer electricity use and -59.7 cubic feet
per year for winter natural gas use. Both results are statistically
significant, with Mann-Kendall p-values < 0.001.

For reference, the ordinary least squares linear trend in summer CDD is also
positive and significant: +3.7 degree days per year (p < 0.001). The trend in
winter HDD is negative and significant: -6.9 degree days per year
(p < 0.001).

Previous research studies (e.g., Alipour et al., 2019; Mukherjee et al.,
2019; Nateghi & Mukherjee, 2017) have shown that linear models may not fully
capture the complex relationships between energy and climate. EPA recognizes
this limitation and notes that ordinary least-squares and Sen's slope linear
regression have been used here for first-order screening purposes only.

## Technical Documentation

```{r}
#| label: td-link
#| echo: false
#| output: asis
technical_documentation_link(meta)
```

## References

<ol class="references">
  <li id="ref-1">EIA (U.S. Energy Information Administration). (2024). <em>Residential Energy Consumption Survey (RECS)</em>. Retrieved December 11, 2024, from [www.eia.gov/consumption/residential/index.php](https://www.eia.gov/consumption/residential/index.php)</li>
  <li id="ref-2">Zamuda, C. D., Bilello, D. E., Carmack, J., Davis, X. J., Efroymson, R. A., Goff, K. M., Hong, T., Karimjee, A., Loughlin, D. H., Upchurch, S., & Voisin, N. (2023). Chapter 5: Energy supply, delivery, and demand. In USGCRP (U.S. Global Change Research Program), <em>Fifth National Climate Assessment</em>. [https://doi.org/10.7930/NCA5.2023.CH5](https://doi.org/10.7930/NCA5.2023.CH5)</li>
  <li id="ref-3">EIA (U.S. Energy Information Administration). (2024). <em>Frequently asked questions (FAQs): What is U.S. electricity generation by energy source?</em> [Data derived from <em>Electric Power Monthly</em> and <em>Electric Power Annual</em>]. [www.eia.gov/tools/faqs/faq.php?id=427&t=3](https://www.eia.gov/tools/faqs/faq.php?id=427&t=3)</li>
  <li id="ref-4">EIA (U.S. Energy Information Administration). (2024). <em>Monthly Energy Review</em>. Retrieved December 5, 2024, from [www.eia.gov/totalenergy/data/monthly](https://www.eia.gov/totalenergy/data/monthly)</li>
  <li id="ref-5">BEA (U.S. Bureau of Economic Analysis). (2024). <em>Population (POPTHM)</em>. FRED, Federal Reserve Bank of St. Louis. Retrieved December 5, 2024, from [https://fred.stlouisfed.org/series/POPTHM](https://fred.stlouisfed.org/series/POPTHM)</li>
  <li id="ref-6">NOAA (National Oceanic and Atmospheric Administration). (2024). <em>National Centers for Environmental Information</em>. Retrieved December 5, 2024, from [www.ncei.noaa.gov](https://www.ncei.noaa.gov)</li>
  <li id="ref-7">EIA (U.S. Energy Information Administration). (2023). <em>Annual energy outlook 2023</em>. [www.eia.gov/outlooks/aeo](https://www.eia.gov/outlooks/aeo)</li>
</ol>

::: {.callout-note appearance="simple" icon=false}
Text on this page is EPA's own published wording (a U.S. Government work, not
subject to domestic copyright), extracted from EPA's source document and
verified paragraph by paragraph against the published page. Data, chart code,
and the full extraction pipeline are maintained in the
[residential-energy-use](https://github.com/climateindicators/residential-energy-use)
repository, which this page reads its data from directly.
:::
